% \documentclass[draft]{report}

\RequirePackage{luatex85}% TeXLive 2017 fix for \geometry
\documentclass{kdp}
\usepackage{geometry}

% \usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
% \usepackage{algorithm}% http://ctan.org/pkg/algorithms
% \usepackage{algpseudocode}
\usepackage{tlatex}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{fancyhdr}
\usepackage{amssymb}
\usepackage{inputenc}
\usepackage{svg}
\usepackage{framed}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}

% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     citecolor=black,
%     filecolor=black,
%     linkcolor=black,
%     urlcolor=black
% }

% !TeX spellcheck = en_GB 

% Configure fancyhdr
\pagestyle{fancy}
\fancyhf{} % Clear default header and footer

% Header settings
\fancyhead[L]{\nouppercase{\leftmark}} % Chapter number and title on the left
% \fancyhead[C]{Center Header}    % Centered header
\fancyhead[R]{\thepage}     % Right-aligned header

% Footer settings
% \fancyfoot[L]{Left Footer}      % Left-aligned footer
% \fancyfoot[C]{Page \thepage}    % Centered footer with page number
% \fancyfoot[R]{Right Footer}     % Right-aligned footer

% java -cp /home/richard/dev/tla2tex/tla2tools.jar  tla2tex.TeX  book.tex 

\lstset { %
    language=C++,
    backgroundcolor=\color{black!5}, % set backgroundcolor
    basicstyle=\footnotesize,% basic font setting
}

\title{\textit{Correct by Design} with TLA+ \\ Early Preview}

% \maketitle



\author{Richard Tang}
\date{\today}
\begin{document}
\maketitle

% \vspace{1cm} % Adjust spacing as needed
% \begin{center}
% \textbf{First Draft}
% \end{center}

\section*{Acknowledgement}

A big thanks to Anthony Giardina for reviewing the content of this book.

\tableofcontents

\part{Introduction}

\include{intro}

\include{primer}

\part{Examples with TLA+}

TLA+ notation is rooted in temporal logic and doesn't share the usual
programming language \textit{look and feel}. Despite the possibly foreign look,
the core language semantics for TLA+ is reasonably concise. This makes it
accessible to anyone with some programming experience, allowing them to learn it
relatively quickly. This section presents a series of TLA+ specification
examples with increasing complexity, designed to gently introduce readers to
this powerful tool.

% \include{blinking}

\include{dining}

\include{scc}

\include{scheduler}

\include{gossip}

\include{sack}

\include{torrent}

\include{raft}

\part{Examples with PlusCal}

PlusCal is a C-like syntax that allows designers to describe their
specifications using a pseudo programming language. An use case for PlusCal is
to specify lockless or wait-free data structures.\\

It is possible to express concurrent data structures in TLA+ native notation. To
express a function that can be concurrently executed, the specification needs to
use a PC-like variable to track the statement a context is currently executing.
This is fairly tedious (and possibly error-prone), making PlusCal a more organic
tool to express these designs. 

\include{spsc}

\include{spmc}

\part{System Modeling}

Specifications described so far have been designed to with finite state space to
allow model checker to explore all states and prove correctness. In this
section, we will relax this and allow infinite state space.\\

This uses TLA+ as a \textit{prototyping} tool. By definition, the model checker
cannot verify infinite sate space. However, for all states it does explore, it
will verify safety properties and highlight all the violations. This enables
designer to quickly iterate on design to flesh out what may or may not work. Per
80/20 rule, the model checker will very quickly identify obvious violations in
the design.\\

Once settled on a design and model checker no longer reports violation in any 
practical amount of time, one can always simplify the specification into finite
space to exhaustively prove correctness. 

\include{lru}

\include{dropbox}

\include{consistent}

\part{Reference}

\include{fairness}

\include{liveness}

\include{general}

\include{lang}

\include{reference}

\end{document}
